\documentclass{beamer}

\input{settings.tex}

\newcommand{\degree}{^{\circ}}     



\title{Frequency response, Bode}
\subtitle{Control Theory, Lecture 5}
\author{by Sergei Savin}
\centering
\date{\mydate}



\begin{document}
\maketitle


\begin{frame}{Content}

\begin{itemize}
\item Laplace and Fourier transforms
\item Laplace and steady state solution
\item Bode plot
\item Bode plot - example
\item Stability margins
\item Code example
\end{itemize}

\end{frame}



\begin{frame}{Frequency response}
	% \framesubtitle{O}
	\begin{flushleft}
		
		\begin{block}{Frequency response}
			Frequency response is a steady-state output of the system, given sinusoidal input.
		\end{block}
	
		\bigskip
	
		Consider a system $Y(s) = G(s)U(s)$. Sinusoidal input $u(t) = \text{sin}(\omega t)$ in time domain translates to $U(s) = \frac{\omega}{\omega^2 + s^2}$ in Laplace domain. So, given a sinusoidal input, the system becomes:
		
		\begin{equation}
			Y(s) = G(s)\frac{\omega}{\omega^2 + s^2}
		\end{equation}
		
	\end{flushleft}
\end{frame}



\begin{frame}{Fraction expansion}
	% \framesubtitle{O}
	\begin{flushleft}
		
		If a transfer function $G(s)$ is a rational fraction, it can be represented as:
		
		\begin{equation}
			G(s) = \frac{n(s)}{(s + p_1)(s + p_2) \ ... \ (s + p_n)}
		\end{equation}		
	
		where $p_i$ are the roots on the denominator - called \emph{poles} of the transfer function.
		
		\bigskip
		
		In many cases (for example when $p_i$ are real and non-repeating), the fraction can be expanded:
		
		\begin{equation*}
			G(s) = \frac{n(s)}{(s + p_1)(s + p_2) \ ... \ (s + p_n)} = \frac{r_1}{s + p_1} + \frac{r_2}{s + p_2} + ... + \frac{r_n}{s + p_n}
		\end{equation*}		
		
		
	\end{flushleft}
\end{frame}



\begin{frame}{Fraction expansion}
	% \framesubtitle{O}
	\begin{flushleft}
		
		We can expand the function $Y(s) = G(s)\frac{\omega}{\omega^2 + s^2}$ in a similar way:
		
		\begin{equation*}
			Y(s) = \frac{r_1}{s + p_1} + \frac{r_2}{s + p_2} + ... + \frac{r_n}{s + p_n} + \frac{\alpha}{s + j\omega} + \frac{\beta}{s - j\omega}
		\end{equation*}		
	
		Laplace function of the form $\frac{r_i}{s + p_i}$ corresponds to the following time function:
		
		\begin{align}
			y(t) = r_i e^{-p_i t}
		\end{align} 
	
		So, for a stable transfer function as time goes to infinity, $r_i e^{-p_i t}$ goes to zero. The only components of the function $Y(s)$ that do not disappear are the last two: $\frac{\alpha}{s + j\omega} + \frac{\beta}{s - j\omega}$.
		
		
	\end{flushleft}
\end{frame}



\begin{frame}{Fraction expansion}
	% \framesubtitle{O}
	\begin{flushleft}
		
		One can show that constants in the expansion $\frac{\alpha}{s + j\omega} + \frac{\beta}{s - j\omega}$ can be found in the form:
		
		\begin{align}
			\alpha = -G(j\omega) g
			\\
			\beta = G(-j\omega) g
		\end{align}
	
		\bigskip
	
		In fact, the analysis of the frequency response will involve analyzing the transfer function $G(j\omega)$.
		
	\end{flushleft}
\end{frame}






\begin{frame}{Laplace and Fourier transforms}
% \framesubtitle{O}
\begin{flushleft}

\begin{itemize}
    \item \emph{Fourier series} can be seen as representing a periodic function as a sum of harmonics (sines and cosines). These sines and cosines can be thought of as forming a basis in a linear space. The coefficients of the series can be thought of as a discrete spectrum of the function.
    
    \item \emph{Fourier transform} gives a continuous spectrum of the function. The "basis" is still made of harmonic functions.
    
    \item \emph{Laplace transform} also gives a continuous spectrum of the function, but in a different basis: the basis is given by complex exponentials. I like to think of this basis as solutions of second order ODEs.
\end{itemize}

\end{flushleft}
\end{frame}




\begin{frame}{Laplace and Fourier transforms}
% \framesubtitle{O}
\begin{flushleft}

Let's compare. Fourier transform:

\begin{equation}
    F(\omega) = \int_{-\infty}^\infty f(t) e^{-2\pi j t \omega} dt, \ \ \omega \in \mathbb{R}
\end{equation}

Laplace transform:

\begin{equation}
    F(s) = \int_0^\infty f(t) e^{-st}dt, \ \ s \in \mathbb{C}
\end{equation}

We can see that Fourier looks like Laplace with purely imaginary number in the exponent.

\end{flushleft}
\end{frame}







\begin{frame}{Laplace and steady state solution}
% \framesubtitle{O}
\begin{flushleft}

From analysing solutions of linear ODEs we know that, given harmonic input (sine, cosine, their combination) "after the transient process is over, the solution approaches a harmonic with the same frequency", but possibly different amplitude and phase.
 
\bigskip

Intuitively we can think of the imaginary part of $s$ as having to do with this frequency response. 

\bigskip
The kernel function of the Laplace transform is $e^{-st}$ with $s = \sigma + j \omega$ being a complex variable. If $\sigma = 0$, the kernel becomes $e^{-j \omega t} = \text{cos}(\omega t) - j \text{sin}(\omega t)$. You can see the similarity with Fourier transform kernel.


\end{flushleft}
\end{frame}





\begin{frame}{Bode plot}
% \framesubtitle{O}
\begin{flushleft}

The first key idea of a Bode plot is substitution of purely complex variable $j \omega$ in place of Laplace variable $s$, which can have non-zero real part.

\bigskip

Given a transfer function $W(s)$, $s = \sigma + j \omega$ we can analyse its behaviour when $\sigma = 0$. We can plot its amplitude $a(\omega) = \left| W(j \omega) \right|$ and its phase $\varphi(\omega) = \text{atan2}( \text{im}(W(j \omega)), \ \text{real}(W(j \omega))  )$.

\bigskip
 
Bode plot is actually two plots, 1) $20 \cdot \text{log}(a(\omega))$ and 2) $\frac{180}{\pi} \varphi(\omega)$. The 20 and log has to do with the vertical axis being in decibels. 

\end{flushleft}
\end{frame}




\begin{frame}{Bode plot - example}
% \framesubtitle{O}
\begin{flushleft}

Consider $W(s) = \frac{1}{1 + s}$. Then $W(j \omega) = \frac{1}{1 + j \omega}$. We can transform it as:

\begin{equation}
    W(j \omega) = \frac{1 - j \omega}{(1 + j \omega)(1 - j \omega)} = 
    \frac{1 - j \omega}{1 + \omega^2}
\end{equation}

Thus we have $\text{real}(W(j \omega)) = \frac{1}{1 + \omega^2}$ and $\text{im}(W(j \omega)) = - \frac{\omega}{1 + \omega^2}$.

\bigskip
 
Bode plot is then given as:

\begin{equation}
   a(\omega) = \sqrt{\frac{1 + \omega^2}{(1 + \omega^2)^2}} = 
   \frac{1}{\sqrt{(1 + \omega^2)}}
\end{equation}
\begin{equation}
   \varphi(\omega) = \text{atan2} \left(-\frac{\omega}{1 + \omega^2}, \ \frac{1}{1 + \omega^2} \right)
\end{equation}

\end{flushleft}
\end{frame}





\begin{frame}{Bode plot - stability margins}
% \framesubtitle{O}
\begin{flushleft}

Before we discuss the use of Bode plot, let us remember that closed-loop transfer function has form (when simple feedback is used):

\begin{equation}
    W(s) = \frac{G(s)}{1 + G(s)}
\end{equation}

Substituting $s \longrightarrow j \omega$ we get:

\begin{equation}
    W(\omega) = \frac{G(j \omega)}{1 + G(j \omega)}
\end{equation}

From this we can see that $W(\omega)$ becomes ill-defined if $G(j \omega) = -1$. Meaning, we want to avoid two things happening simultaneously: the amplitude of $G(j \omega)$ being equal to 1, and its phase (argument) being equal to $180\degree$ (remember, phase of $0\degree$ is pure positive real number, phase of $90\degree$ is pure positive imaginary number, $180\degree$ is pure negative real number, etc.).

\end{flushleft}
\end{frame}





\begin{frame}{Stability margins - graphical example}
% \framesubtitle{O}
\begin{flushleft}

Let's check an illustration:

\bigskip

\centerline{\textcolor{black}{\qrcode[height=1.6in]{https://www.electrical4u.com/bode-plot-gain-margin-phase-margin/}}}


\end{flushleft}
\end{frame}




\begin{frame}{Code example}
% \framesubtitle{O}
\begin{flushleft}

Check the colab notebook based on the example above for an illustration of how the Bode plot can be made by hand or via scipy signal library.

\bigskip

\centerline{\textcolor{black}{\qrcode[height=1.6in]{https://github.com/SergeiSa/Control-Theory-Slides-Spring-2022/blob/main/ColabNotebooks/lecture_Bode.ipynb}}}


\end{flushleft}
\end{frame}




\begin{frame}{Read more}

\begin{itemize}
\item \bref{https://youtu.be/_eh1conN6YM}{Control System Lectures - Bode Plots, Introduction}

\item \bref{https://global.oup.com/us/companion.websites/fdscontent/uscompanion/us/static/companion.websites/9780199339136/Appendices/Appendix_F.pdf}{Oxford University Press. s-Domain analysis: poles, zeros, and Bode plots}


\end{itemize}

\end{frame}




\myqrframe

\end{document}
